### 2022-10-15 AISTATS2023/paper.tex
ready for submission

\begin{abstract}
  Accumulated Local Effects (ALE) is a popular explainable AI method that quantifies how a feature influences the decisions of a model, handling well feature correlations. In case of complex interactions between features, instance-level feature effects may deviate from the ALE curve. It is crucial to quantify this deviation, namely, the uncertainty of the global effect. In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and its uncertainty. We show that UALE quantifies uncertainty effectively, even in case of correlated features. We also note that as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting). The average effect and the uncertainty are computed from the instances that lie in each bin. We formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints. Based on this, we propose a method to determine the optimal intervals, balancing the estimation bias and variance. We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods and (b) the effectiveness of UALE's appropriate bin splitting for accurately approximating of the average effect and its uncertainty.
\end{abstract}

@@

In case of complex interactions between features, the instance-level (local) feature effects may exhibit significant deviation, aka heterogeneity, from the aggregated (global) explanation, a phenomenon known as aggregation bias~\citep{mehrabi2021survey}. We call this deviation \emph{the uncertainty of the global effect}\footnote{In the paper, the \emph{uncertainty of the global effect} and the \emph{heterogeneity of instance-level effects} are used interchangeably and refer to the deviation of instance-level (local) effects from the averaged feature effect curve (global).}. Aggregation bias leads to ambiguities. For example, a feature with zero average effect may indicate (a) no effect on the output or (b) highly positive effect for some instances and highly negative for some others. As a result, there is a need for FE methods to quantify the deviation of instance-level explanations, in addition to the average effect. In the case of PDP, such deviation is visualized via the Individual Conditional Explanations (ICE)~\citep{goldstein2015peeking}. ICE plots, however, have the same limitations as PDP in case of correlated features. No method to quantify the deviation of instance-level effects has been proposed for ALE.

@@

\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{concept_figure/fig-1.pdf}
  \caption{Illustration of UALE feature effect plot.}
  \label{fig:UALE-figure}
\end{figure}

@@

UALE defines the effect as a pair that consists of the average effect and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}), and the accumulated uncertainty, as defined in Eq.~(\ref{eq:ALE_acc_unc}). Figure~\ref{fig:UALE-figure} illustrates a toy example, where \(x_1 \sim \mathcal{U}(0,1)\) and \(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and the black-box function is \(f(\xb)= \sin(2 \pi x_1) + x_1x_2\).

@@

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
where \(k_x\) is the index of the bin such that
\( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
Eq.~(\ref{eq:ALE_accumulated_var}) are piecewise-linear approximation
of Eq.~(\ref{eq:ALE}) and Eq.~(\ref{eq:ALE_acc_unc}).

@@

%
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
%
In the Appendix~\ref{sec:proof-1}, we show that
\(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|X_s=z}
  \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
  z_1} \). In Theorem~\ref{sec:theorem-1} we prove that in the general
case, \(\sigma_*^2(z_1, z_2) \geq \sigma^2(z_1, z_2)\), so
\(\hat{\sigma}^2(z_1, z_2)\) is an overestimation of the bin
uncertainty.

@@

\begin{proof}
The proof is at Appendix~\ref{sec:proof-of-3-1}.
\end{proof}

@@

\begin{corollary}
  \label{sec:coroll}
  A bin-splitting \(\mathcal{Z}\) that minimizes the accumulated error
  if and only if it also minimizes
  \(\sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k \)
\end{corollary}

\begin{proof}
  ~\label{sec:coroll-1} The proof is based on the observation that
  \(\sum_{k=1}^K \sigma^2(z_{k-1}, z_k) \Delta z_k = \sigma^2(z_0,
  z_K) (z_K - z_0)\) which is independent of the bin-splitting. A more
  detailed proof is provided in the Appendix~\ref{sec:proof-of-coroll}.
 \end{proof}

@@

To define the solution, we use two indexes. The index
\(i \in \{0, \ldots, K_{max}\}\) denotes the points of the
partitioning \((z_i)\) and the index \(j \in \{0, \ldots, K_{max}\} \)
denotes the \(j\)-th multiple of the minimum step, i.e.,
\(x_j = x_{s,min} + j \cdot \Delta x_{min}\). The recursive cost
function \(T(i,j)\) computes the cost of setting \(z_i\) to \(x_j\):
\begin{equation}
  \label{eq:recursive_cost}
  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
\end{equation}
%
% where \(\mathcal{T}(0,j)\) equals to zero if \(j=0\) and \(\infty\) in
% any other case.
The term \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin with
limits \([x_l, x_j)\):

\begin{equation}
  \label{eq:cost_step}
  \mathcal{B}(x_l, x_j) = \begin{cases}
                            \infty, & \text{if $x_j > x_l$ or \(|\mathcal{S}_{(x_j, x_l)}| < N\)}\\
                            0, & \text{if $x_j = x_l$}\\
                            \hat{\sigma}^2(x_j, x_l), &\text{if $x_j \leq x_l$}
  \end{cases}
\end{equation}

The optimal solution is given by solving
\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of the sequence of
steps.

% \begin{algorithm}
%   \end{algorithm}

@@

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_1(\mathbf{x}) < 1$}\\
                    \alpha f_2(\mathbf{x}) &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_2(\mathbf{x}) = x_1x_3\). The part \(f_1\) is a linear function of
the two correlated features, \(x_1, x_2\), and \(f_2\) is an
interaction term between the two non-correlated, \(x_1, x_3\).  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weigths
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) and
equal weights (\(a_1=a_2\)).

In the general case, evaluating feature effect methods is a
challenging task, because there is no unique ground-truth feature
effect and uncertainty and each method defines the effect in a
different way. To deal with this, we exploit two characteristics of
the above set-up. First, in \(p(\xb)\) it holds that
\(x_1=x_2\). Therefore, knowing \(x_1\) or \(x_2\), we know the
closed-form of \(f(\xb)\), too. For example, if \(a_1=a_2=1\) and
\( 0 \leq x_1 < \frac{1}{4}\), then \(f_1(\xb) < \frac{1}{2}\), so
\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \), i.e., the first branch of
Eq.~\ref{eq:synth-ex-1-function}. Afterwards, we can determine the
ground-truth. In the example above, if \(a=0\), then
\(f(\xb) = x_1 + x_2\) and therefore the ground-truth effect of
\(x_1\) is \(f^{\mathcal{GT}}(x_1) = x_1\) in the corresponding
interval, i.e. \(x_1 \in [0, \frac{1}{4})\). For the uncertainty, we
use the fact that under no-interactions, \(a=0\), the uncertainty must
be zero (no heteregoneity). We discuss separately the case with
\(a>0\). Finally, we also demonstrate that UALE's approximation is
accurate, due to the bin splitting that we propose.

@@

Here, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\) (equal
weights). We examine only the effect of feature \(x_1\), because it is
exactly the same as with \(x_2\). Based on the discussion above, the
ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) in
\([0, \frac{1}{4}]\), \(-x_1\) in \([\frac{1}{4}, \frac{1}{2}]\) and
\(0\) in \([\frac{1}{2}, 1]\). Because \(x_1\) does not interact with
any other feature, the uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
wrong and ICE plots show heterogeneous effects. In contrast, UALE
models correctly both the average effect and the absence of
uncertainty. Finally, we observe that UALE's bin-splitting leads (a)
to an accurate uncertainty estimation (near-zero cumulative bin-error
\(\mathcal{E}_{\mathcal{Z}}\)) and (b) an easier interpretation,
creating three wide bins, i.e.
\([0, \frac{1}{4}), [\frac{1}{4}, \frac{1}{2}), [\frac{1}{2}, 1)\),
that correspond to the piecewise-linear regions of the effect.

@@

Here, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
\(a_2= \frac{1}{2}\) (non-equal weights). The non-equal weights have
implications at both the gradient and the interval of the piece-wise
linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
\([0, \frac{1}{5}]\), \(-2x_1\) in \([\frac{1}{5}, \frac{2}{5}]\) and
\(0\) in \([\frac{2}{5}, 1]\). As before, the ground-truth uncertainty
is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
interact with any other feature. In Figure~\ref{fig:ex-synth-1-2}, we
observe that PDP estimation is opposite to the ground-truth effect,
i.e.~negative in the region \([0, \frac{1}{5})\), positive in
\([\frac{1}{5}, \frac{2}{5})\), and the ICE erroneously implies the
existence of heterogeneous effects. As before, UALE quantifies
correctly the ground truth effect, the zero-uncertainty and partitions
the \(x_1\) domain into wide bins that facilitate the interpretation
and create a zero cummulative bin-error \(\mathcal{E}_{\mathcal{Z}}\)
approximation.

@@

We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) UALE's partitioning, and
with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with \(K\)
bins. The evaluation is based on the Mean Absolute Error (MAE) of the
bin effect \(\mu\) and of the uncertainty \(\sigma\) accross bins,
i.e.,

@@

We compare UALE vs fixed-size approximation, when (a) \(f\) is
piecewise-linear and (b) \(f\) is non-linear. In all examples, we
execute \(t = 30\) independent runs, using each time \(N=500\)
different samples, and we report the mean values of the metrics.

\paragraph{Piecewise-Linear Function.}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
  \caption{Bin-Splitting, piecewise-linear function: UALE's approximation
    (Top-Left). UALE vs fixed-size approximations in terms of:
    \(\mathcal{L}^{\mu}\) (Top-Right), \(\mathcal{L}^{\sigma}\)
    (Bottom-Left), \(\mathcal{L}^{\rho}\) (Bottom-Right).}
  \label{fig:ex-synth-2-1}
\end{figure}

@@

Here, we aim at demonstrating the usefuleness of uncertainty
quantification and the advantages of UALE's approximation, on the
real-world California Housing dataset~\citep{pace1997sparse}.

@@

Below, we illustrate the feature effect for three features: latitude
\(x_2\), population \(x_6\) and median income \(x_8\). The particular
features cover the main FE cases, e.g.~positive/negative trend and
linear/non-linear curve, and are therefore appropriate for
illustration purposes. Results for all features, along with in-depth
information about the preprocessing, training and evaluation parts are
provided in the Appendix.

@@

We have introduced UALE, a method that extends ALE by quantifying the uncertainty of the global feature effect. To do that, UALE uses the standard deviation of instance-level effects. We showed that UALE models the uncertainty of feature effect more accurately than existing methods, such as ICE plots, especially in cases where there are dependencies between features. Providing an unbiased estimate of the uncertainty of the feature effect, however, is not trivial. In the case of ALE and the approximation of the average global effect, splitting the feature domain in K equal-width bins is sufficient. For estimating the uncertainty, however, we proved that this approach would introduce bias. We overcame this limitation by proposing an algorithm for automatically determining the optimal variable-size partitioning considering the instance-level effects. We show, through both synthetic and real-world examples, that UALE is an effective method for modeling both the average effect and the uncertainty.

There are certain limitations and open issues in the proposed approach. First, finding the optimal bin partitioning can be computationally expensive, if applied directly to ALE. By using DALE's approximation this limitation is resolved, however this is directly applicable only for differentiable models. Furthermore, UALE quantifies only the standard deviation of instance-level effects, providing limited intuition about the sources of the uncertainty. Future work should therefore aim at developing methods for identifying which features and to what degree contribute to the heterogeneity of instance-level effects.

\clearpage
\subsubsection*{Acknowledgments}
All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support. 
To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.

>>>>

\begin{abstract}
  Accumulated Local Effects (ALE) is a popular explainable AI method that quantifies how a feature influences the decisions of a model, handling well feature correlations. In case of complex interactions between features, instance-level feature effects may deviate from the ALE curve. It is crucial to quantify this deviation, namely, the uncertainty of the global effect. In this work, we define Uncertainty-aware ALE (UALE) to quantify and visualize, on a single plot, both the average effect and its uncertainty. We also note that as in ALE, UALE's approximation requires partitioning the feature domain into non-overlapping intervals (bin-splitting). The average effect and the uncertainty are computed from the instances that lie in each bin. We formally prove that to achieve an unbiased approximation of the uncertainty in each bin, bin-splitting must follow specific constraints. Based on this, we propose a method to determine the optimal intervals, balancing the estimation bias and variance. We demonstrate, through synthetic and real datasets, (a) the advantages of modeling the uncertainty with UALE compared to alternative methods, especially in cases of correlated features and (b) the effectiveness of UALE's appropriate bin splitting for accurately approximating of the average effect and its uncertainty.
\end{abstract}

@@

In case of complex interactions between features, the instance-level (local) feature effects may exhibit significant deviation, aka heterogeneity, from the aggregated (global) explanation, a phenomenon known as aggregation bias~\citep{mehrabi2021survey}. We call this deviation \emph{the uncertainty of the global effect}\footnote{In the paper, the \emph{uncertainty of the global effect} and the \emph{heterogeneity of instance-level effects} are used interchangeably and refer to the deviation of instance-level (local) effects from the averaged feature effect curve (global).}. Aggregation bias leads to ambiguities. For example, a feature with zero average effect may indicate (a) no effect on the output or (b) highly positive effect for some instances and highly negative for some others. As a result, there is a need for FE methods to quantify the deviation of instance-level explanations, in addition to the average effect. In the case of PDP, such deviation is visualized via the Individual Conditional Explanations (ICE)~\citep{goldstein2015peeking}. ICE plots, however, have the same limitations as PDP~\citep{baniecki2021fooling} in case of correlated features. No method to quantify the deviation of instance-level effects has been proposed for ALE.

@@

\begin{figure}
  \centering
  \includegraphics[width=.4\textwidth]{concept_figure/fig-1.pdf}
  \caption{Illustration of UALE feature effect plot on a toy example,
    where \(x_1 \sim \mathcal{U}(0,1)\) and
    \(x_2 \sim \mathcal{N}(0,\sigma^2_2)\) and
    \(f(\xb)= f_0(\xb) + x_1x_2\), where \(f_0(\xb)\) when
    \(x_1 < \frac{1}{5}\) and zero otherwise.}
  \label{fig:UALE-figure}
\end{figure}

@@

UALE defines the effect as a pair that consists of the average effect
and the uncertainty \((\mu(z), \sigma(z))\). For visualizing it, we
use the accumulated average effect, as defined in Eq.~(\ref{eq:ALE}),
and the accumulated uncertainty, as defined in
Eq.~(\ref{eq:ALE_acc_unc}).

@@

\begin{equation}
  \label{eq:ALE_accumulated_var}
  \tilde{f}^{\mathtt{ALE}}_{\mathcal{Z}, \sigma}(x_s) =  \sum_{k=1}^{k_x} \sigma(z_{k-1}, z_k) (z_k - z_{k-1})
\end{equation}
%
where \(k_x\) is the index of the bin such that
\( z_{k_x - 1} \leq x_s < z_{k_x}\). Eq.~(\ref{eq:ALE_2}) and
Eq.~(\ref{eq:ALE_accumulated_var}) are piecewise linear approximation
of Eq.~(\ref{eq:ALE}) and Eq.~(\ref{eq:ALE_acc_unc}).

@@

%
\begin{equation}
  \label{eq:var_bin_approx}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2
\end{equation}
%
In the Appendix %~\ref{sec:proof-1}
, we show that
\(\hat{\sigma}^2(z_1, z_2)\) is an unbiased estimator of
\(\sigma^2_*(z_1, z_2) = \frac{\int_{z_1}^{z_2} \mathbb{E}_{X_c|X_s=z}
  \left [ (f^s(z, X_c) - \mu(z_1, z_2) )^2 \right] \partial z}{z_2 -
  z_1} \). In Theorem~\ref{sec:theorem-1} we prove that in the general
case, \(\sigma_*^2(z_1, z_2) \geq \sigma^2(z_1, z_2)\), so
\(\hat{\sigma}^2(z_1, z_2)\) is an overestimation of the bin
uncertainty.

@@

\begin{proof}
The proof is at Appendix.%~\ref{sec:proof-of-3-1}.
\end{proof}

@@

\begin{corollary}
  \label{sec:coroll}
  If a bin-splitting \(\mathcal{Z}\) minimizes the accumulated error,
  then it also minimizes
  \(\sum_{k=1}^K\sigma_*^2(z_1, z_2) \Delta z_k \)
\end{corollary}

\begin{proof}
  ~\label{sec:coroll-1} The proof is based on the observation that
  \(\sum_{k=1}^K \sigma^2(z_{k-1}, z_k) \Delta z_k = \sigma^2(z_0,
  z_K) (z_K - z_0)\) which is independent of the bin-splitting. A more
  detailed proof is provided in the Appendix.%~\ref{sec:proof-of-coroll}.
 \end{proof}

@@

To define the solution, we use two indexes. The index
\(i \in \{0, \ldots, K_{max}\}\) denotes the points of the
partitioning \((z_i)\) and the index \(j \in \{0, \ldots, K_{max}\} \)
denotes the \(j\)-th multiple of the minimum step, i.e.,
\(x_j = x_{s,min} + j \cdot \Delta x_{min}\). The recursive cost
function \(T(i,j)\) computes the cost of setting \(z_i\) to \(x_j\):
\begin{equation}
  \label{eq:recursive_cost}
  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(l, j) \right ]
\end{equation}
%
% where \(\mathcal{T}(0,j)\) equals to zero if \(j=0\) and \(\infty\) in
% any other case.
The term \(\mathcal{B}(l, j)\) denotes the cost of creating a bin with
limits \([x_l, x_j)\). \(\mathcal{B}\) can be custom-based. In our
case, following Eq.~(\ref{eq:opt}), we set it to
\(\tau_k \hat{\sigma}^2(x_l, x_j) (x_j-x_l)\) if the bin is valid,
i.e., it has enough points \(|\mathcal{S}_k| \geq
N_{\mathtt{PPB}}\). The
optimal solution is given by solving
\(\mathcal{L} = \mathcal{T}(K_{max}, K_{max})\) and keeping track of
the sequence of steps.

Note that in case the instance level effects are computed using the
DALE method then there is no need to reevaluate \(f\) for each
partitioning. The optimization algorithm simply reallocates the
precomputed instance-level effects at each bin. Therefore, setting the
maximum number \(K_{max}\) up to (roughly) \(100\) bins, the method
finds the optimal partitioning in a couple of seconds. Further details
and evaluation are provided in the Appendix.

@@

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $f_1(\mathbf{x}) < \frac{1}{2}$ }\\
                    \frac{1}{2} - f_1(\mathbf{x}) + \alpha f_2(\mathbf{x}) & \text{if $\frac{1}{2} \leq f_1(\mathbf{x}) < 1$}\\
                    \alpha f_2(\mathbf{x}) &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_1(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_2(\mathbf{x}) = x_1x_3\). The part \(f_1\) is a linear function of
the two correlated features, \(x_1, x_2\), and \(f_2\) is an
interaction term between the two non-correlated, \(x_1, x_3\).  We
evaluate the effect computed by UALE and PDP-ICE in three cases; (a)
without interaction (\(\alpha=0\)) and equal weights (\(a_1=a_2\)),
(b) without interaction (\(\alpha=0\)) and different weights
(\( a_1 \neq a_2 \)) and (c) with interaction (\(\alpha > 0\)) and
equal weights (\(a_1=a_2\)).

In the general case, evaluating feature effect methods is a
challenging task, because there is no unique ground-truth feature
effect and uncertainty and each method defines the effect in a
different way. To deal with this, we exploit two characteristics of
the above set-up. First, in \(p(\xb)\) it holds that
\(x_1=x_2\). Therefore, knowing \(x_1\) or \(x_2\), we know the
closed-form of \(f(\xb)\), too. For example, if \(a_1=a_2=1\) and
\( 0 \leq x_1 < \frac{1}{4}\), then \(f_1(\xb) < \frac{1}{2}\), so
\(f(\xb) = f_1(\xb) + \alpha f_2(\xb) \), i.e., the first branch of
Eq.~\ref{eq:synth-ex-1-function}. Afterwards, we can determine the
ground-truth. In the example above, if \(a=0\), then
\(f(\xb) = x_1 + x_2\) and therefore the ground-truth effect of
\(x_1\) is \(f^{\mathcal{GT}}(x_1) = x_1\) in the corresponding
interval, i.e. \(x_1 \in [0, \frac{1}{4})\). For the uncertainty, we
use the fact that under no-interactions, \(a=0\), the uncertainty must
be zero (no heterogeneity). We discuss separately the case with
\(a>0\). Finally, we also demonstrate that UALE's approximation is
accurate, due to the bin splitting that we propose.

@@

Here, \(\alpha=0\) (no interaction) and the \(a_1=a_2=1\) (equal
weights). We examine only the effect of feature \(x_1\), because it is
exactly the same as with \(x_2\). Based on the discussion above, the
ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\) in
\([0, \frac{1}{4}]\), \(-x_1\) in \([\frac{1}{4}, \frac{1}{2}]\) and
\(0\) in \([\frac{1}{2}, 1]\). Because \(x_1\) does not interact with
any other feature, the uncertainty is
\(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP effect is
wrong and ICE plots show heterogeneous effects. In contrast, UALE
models correctly both the average effect and the absence of
uncertainty. Finally, we observe that UALE's bin-splitting leads (a)
to an accurate uncertainty estimation (near-zero cumulative bin-error
\(\mathcal{E}_{\mathcal{Z}}\)) and (b) an easier interpretation,
creating three wide bins, i.e.
\([0, \frac{1}{4}), [\frac{1}{4}, \frac{1}{2}), [\frac{1}{2}, 1)\),
that correspond to the piecewise linear regions of the effect.

@@

Here, \(\alpha=0\) (no interaction), \(a_1 = 2\) and
\(a_2= \frac{1}{2}\) (non-equal weights). The non-equal weights have
implications at both the gradient and the interval of the piece-wise
linear regions, i.e., \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(2x_1\) in
\([0, \frac{1}{5}]\), \(-2x_1\) in \([\frac{1}{5}, \frac{2}{5}]\) and
\(0\) in \([\frac{2}{5}, 1]\). As before, the ground-truth uncertainty
is \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\) because \(x_1\) does not
interact with any other feature. In Figure~\ref{fig:ex-synth-1-2}, we
observe that PDP estimation is opposite to the ground-truth effect,
i.e.~negative in the region \([0, \frac{1}{5})\), positive in
\([\frac{1}{5}, \frac{2}{5})\), and the ICE erroneously implies the
existence of heterogeneous effects. As before, UALE quantifies
correctly the ground truth effect, the zero-uncertainty and partitions
the \(x_1\) domain into wide bins that facilitate the interpretation
and create a zero cumulative bin-error \(\mathcal{E}_{\mathcal{Z}}\)
approximation.

@@

We sample from \(p(\mathbf{x}) = p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\) and
\(x_2 \sim \mathcal{N}(x_1, \sigma_2^2=0.5)\). We denote as
\(\mathcal{Z^*} = \{z^*_0, \cdots, z^*_K\}\) UALE's partitioning, and
with \(\mathcal{Z^{\mathtt{K}}}\) the fixed-size splitting with \(K\)
bins. The evaluation is based on the Mean Absolute Error (MAE) of the
bin effect \(\mu\) and of the uncertainty \(\sigma\) across bins,
i.e.,

@@

We compare UALE vs fixed-size approximation, when (a) \(f\) is
piecewise linear and (b) \(f\) is non-linear. In all examples, we
execute \(t = 30\) independent runs, using each time \(N=500\)
different samples, and we report the mean values of the metrics.

\paragraph{Piecewise Linear Function.}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_1.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_2.pdf}\\
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_3.pdf}
  \includegraphics[width=.23\textwidth]{example_bin_splitting_1/fig_4.pdf}
  \caption{Bin-Splitting, piecewise linear function: UALE's approximation
    (Top-Left). UALE vs fixed-size approximations in terms of:
    \(\mathcal{L}^{\mu}\) (Top-Right), \(\mathcal{L}^{\sigma}\)
    (Bottom-Left), \(\mathcal{L}^{\rho}\) (Bottom-Right).}
  \label{fig:ex-synth-2-1}
\end{figure}

@@

Here, we aim at demonstrating the usefulness of uncertainty
quantification and the advantages of UALE's approximation, on the
real-world California Housing dataset~\citep{pace1997sparse}.

@@

Below, we illustrate the feature effect for three features: latitude
\(x_2\), population \(x_6\) and median income \(x_8\). The particular
features cover the main FE cases, e.g.~positive/negative trend and
linear/non-linear curve, and are therefore appropriate for
illustration purposes. Results for all features, along with in-depth
information about the reprocessing, training and evaluation parts are
provided in the Appendix.

@@

In this work we presented the global feature effect method UALE, that
extends ALE to quantify the uncertainty of the feature effect. We
showed that UALE models such uncertainty more accurately than existing
methods, such as ICE plots, especially in cases of highly correlated
features, where that latest fail. Providing an unbiased estimate of
this uncertainty, however, is not trivial. In the case of ALE, for the
approximation of the average effect, splitting the feature domain in
fixed-width bins is sufficient. For estimating the uncertainty,
though, we proved that this approach would introduce bias. To overcame
this limitation, we proposed a method to automatically determine the
optimal variable-width bin-splitting considering the underlying
instance-level effects. We demonstrated, through both synthetic and
real-world examples, that UALE is an effective method to model both
the average effect and the uncertainty.

Future work involves exploring other summary statistics than standard
deviation to reveal further uncertainty aspects. Also, we would like
to study ways to quantify the contribution of each feature in the
uncertainty. Finally, we would like to explore alternatives to
implement UALE's bin splitting (exhaustive vs approximate computation)
to deal with data-intensive cases.

\clearpage
% \subsubsection*{Acknowledgments}
% All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support. 
% To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.
